---
title: "Read in the data from the skin rash samples"
author: "Daniel"
date: "18.05.2020"
output: html_document
---

In this script we read the output from the pre-processing/segmentation pipeline and build a SingleCellExperiment object (https://bioconductor.org/packages/release/bioc/html/SingleCellExperiment.html)

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries

First, we will load the libraries needed for this part of the analysis.

```{r load-libraries, message=FALSE}
library(data.table)
library(S4Vectors)
library(SingleCellExperiment)
library(LSD)
```

# Read the data

```{r load data}
# load the single cell data
cells <- as.data.frame(fread(file = "~/bbvolume/Data/Dermatology/covid_skin_rash/processed/cpout/cell.csv",stringsAsFactors = FALSE))

#nuclei <- as.data.frame(fread(file = "~/bbvolume/Data/Melanoma/12plex/protein/cpout/nuclei.csv",stringsAsFactors = FALSE))

# load the image level metadata
image_mat <- as.data.frame(read.csv(file = "~/bbvolume/Data/Dermatology/covid_skin_rash/processed/cpout/Image.csv",stringsAsFactors = FALSE))

# load the panel information
panel_mat <- read.csv(file = "~/bbvolume/Data/Dermatology/covid_skin_rash/Skin_panel.csv", sep= ",",  stringsAsFactors = FALSE )

# get an example file that contains the channel order
tags <- read.csv( "~/bbvolume/Data/Dermatology/covid_skin_rash/processed/tiffs/20200506_DS_Derma_COV_1_2_s1_p12_r1_a1_ac_full.csv", header = FALSE)

# load acqusition meta data
acquisition_meta <- read.csv(file = "~/bbvolume/Data/Dermatology/covid_skin_rash/processed/cpout/acquisition_metadata.csv", stringsAsFactors = FALSE)
```

# generate the counts data frame

```{r get count data }
cur_counts <- cells[,grepl("Intensity_MeanIntensity_FullStackFiltered",colnames(cells))]
```

# get the scaling factor
the single cell data needs to be multiplied with the scaling factor (16 bit)
```{r scale the count data}
cur_counts <- cur_counts * image_mat$Scaling_FullStack[1]

# to order the channels according to channel number
channelNumber <- as.numeric(sub("^.*_c", "", colnames(cur_counts)))

cur_counts <- cur_counts[,order(channelNumber,decreasing = FALSE)]
```

# prepare the cell-level metadata
this data frame contains the metadata for ever single cell and will later on be the colData in the single cell experiment object

the metadata will also have an entry called "Parent_nuclei" which holds information to the ObjectNumber of the Nuclei that the cell was derived from.
due to the down-scaling of the images some nuclei are lost and thus some cells do not have a Parent_nuclei

```{r column level metadata}

cell_meta <- DataFrame(CellNumber = cells$ObjectNumber,
                       ImageNumber = cells$ImageNumber,
                       Center_X = cells$Location_Center_X,
                       Center_Y = cells$Location_Center_Y,
                       Area = cells$AreaShape_Area,
                       MajorAxisLength = cells$AreaShape_MajorAxisLength,
                       MinorAxisLength = cells$AreaShape_MinorAxisLength,
                       NumberOfNeighbors = cells$Neighbors_NumberOfNeighbors_8)


# add a unique cellID to each cell consisting of "dataset"+"ImageNumber"+"ObjectNumber"
cell_meta$cellID <- paste0(cell_meta$ImageNumber, "_",cell_meta$CellNumber)

rownames(cell_meta) <- cell_meta$cellID
```

# prepare the row-level metadata (panel/marker informatino)
here we prepare all the metadata for the rows in the single cell experiment object (rowData)

```{r row level metadata}
# the channel numbers are the rownumbers in the "tags" file that we create above
tags$channel <- as.numeric(rownames(tags))
colnames(tags) <- c("Metal.Tag","channel")

# include the channel information in the panel metadata (panel_mat)
panel_mat <- merge(panel_mat,tags,by="Metal.Tag")

# now we order the panel metadata by channel. therefore we first modify the column names
panel_mat <- panel_mat[order(panel_mat$channel,decreasing = FALSE),]

# we also revise the nomenclature of the clean targets to not contain special characters like "-" etc
panel_mat$clean_target
clean_target <- c("HistoneH3", "SMA","CD7", "Filaggrin", "HLADR","CD370", "CD27", "CD16", "CD163", "Langerin", "CD11b", "CD68", "CD31",
                  "CD3", "DC_LAMP", "CD11c", "CD1c", "CD69", "CD134", "GrzB", "CD45RA", "STING", "CD45RO", "FoxP3", "CD20", "CLA",
                  "CD8", "CD57", "Ki67", "DP2","CD40", "CD4", "CD14", "E-cad",  "CD303", "CD206", "DNA1", "DNA2", "MPO")
panel_mat$clean_target <- clean_target

rownames(panel_mat) <- panel_mat$clean_target
```

# create the single cell experiment object

```{r create single cell object}
# create the SCE object
sce <- SingleCellExperiment(assays = list(counts = t(cur_counts)))

# Set marker name as rownames and cellID as colnames
rownames(sce) <- rownames(panel_mat)
colnames(sce) <- rownames(cell_meta)

# add the column and row metadata
colData(sce) <- cell_meta
rowData(sce) <- panel_mat

# asinh transformed counts as well as add the nuclear count data
assay(sce, "asinh") <- asinh(counts(sce))
```


## one of the control normal samples was wrongly assigned. It was given the name "NHS_4" but should be "NHS_1". We therefore have "NHS_4" duplicated. We will therefore rename it here manually. The error also becomes visible from the "FileName_CellImage" slot in the "image_mat" object.

```{r add sample information}
# first we fix the broken filename
image_mat[which(image_mat$FileName_CellImage == "20200506_DS_Derma_NHS_1_3_s0_p6_r2_a2_ac_ilastik_s2_Probabilities_mask.tiff"),]$Metadata_Description <- "NHS_1"

image_mat[,c("ImageNumber","FileName_CellImage","Metadata_Description")]

# now we add the sample information to the single cell experiment object
sce$sample <- image_mat$Metadata_Description[sce$ImageNumber]

sce$sample_type <- sapply(as.vector(sce$sample),function(x){strsplit(x,"_",)[[1]][1]})
sce$sample_number <- sapply(as.vector(sce$sample),function(x){strsplit(x,"_",)[[1]][2]})
```


# save the sce object

```{r save sce object}
saveRDS(sce,file = "~/bbvolume/Data/Skin_rash/Rout/data/sce_protein.rds")
```
