---
title: "Plots for manuscript"
author: "daniels"
date: "7 9 2020"
output: html_document
---

This script generates the plots for figures figures 2, 3 and supplementary 3 in Mitamura, Schulz et al.
Also generates some additional plots.

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(echo = TRUE)
```

## load packages

```{r load packages, echo=FALSE, message = FALSE}
#sapply(list.files("~/bbvolume/Git/imcRtools/R/", full.names = TRUE), source)
#sapply(list.files("~/bbvolume/Git/ZTMA256_protein_data/code/helper_functions/", full.names = TRUE), source)

library(LSD)
library(SingleCellExperiment)
library(ggplot2)
library(scater)
library(viridis)
library(Rphenograph)
library(igraph)
library(CATALYST)
library(data.table)
library(tidyverse)
library(dittoSeq)
library(tidyverse)
library(cowplot)
library(ComplexHeatmap)
library(ggpubr)
library(circlize)
```


```{r helper functions }
plotCellFracGroups <- function(x,CellClass,color_by) {
  # check if x is SingleCellExperiment
  
  if (is.null(CellClass)) {
    stop("Provide Cell class from column metadata")
  }


  # check if selected variables exist
  entries <- colnames(colData(x))
  if (! is.null(CellClass)) {
    if (! CellClass %in% entries) {
      stop("The entry for CellClass is not a colData slot of the object.")
    }
  }

  # check if color_by does not exceed the number of images
  if (! is.null(color_by)) {
    if (! color_by %in% entries) {
      stop("The entry for color_by is not a colData slot of the object.")
    }
    if (length(unique(colData(x)[,color_by])) > length(unique(x$ImageNumber))) {
      stop("Number of colors selected not supported")
      }
  }


  # create the data for plotting
  cur_df <- data.frame( "ImageNumber" = as.factor(colData(x)[,"ImageNumber"]))

  cur_df$CellClass <- as.factor(colData(x)[,CellClass])



    cur_df$color_by <- as.factor(colData(x)[,color_by])

    sum_df <- reshape2::dcast(cur_df, ImageNumber + color_by ~ CellClass, length)

    # calculate the fractions for each cell type if color_by is not given
    sum_df <- cbind(sum_df[c("ImageNumber", "color_by")],sum_df[-which(colnames(sum_df) %in% c("ImageNumber", "color_by"))]/rowSums(sum_df[-which(colnames(sum_df) %in% c("ImageNumber", "color_by"))]))

    # melt data for plotting
    frac_df <- reshape2::melt(sum_df,id.vars = c("ImageNumber", "color_by"),variable.name = CellClass)

    colnames(frac_df) <- c("ImageNumber", color_by , CellClass , "value")

    # plot data
    ggplot(frac_df) +
      geom_boxplot(aes_string(x=CellClass,y="value", fill=color_by),position=position_dodge(width=1)) +
      geom_jitter(aes_string(x=CellClass,y="value", fill=color_by),position = position_dodge(width = 1),lwd=0.4) +
      theme_bw()+
      theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
      ylab("Fraction of celltype per Image")


}

scatter_x_y <- function(input_data,x,y,imagenumber,xlim,ylim, select_assay ="asinh"){
  if (is.null(select_assay)) {
    heatscatter(as.vector(t(assay(input_data[rownames(input_data) == x,input_data$ImageNumber == imagenumber],"asinh"))),as.vector(t(assay(input_data[rownames(input_data) == y,input_data$ImageNumber == imagenumber],"asinh"))),
                xlab = paste("Imagenumber:",imagenumber , "Marker:" , x, sep = " "),
                ylab = y,
                xlim = xlim,
                ylim = ylim,
                main=(paste("Image number:", imagenumber,sep =" ")),
                cor = TRUE,
                method = "spearman")
  } else {heatscatter(as.vector(t(assay(input_data[rownames(input_data) == x,input_data$ImageNumber == imagenumber],select_assay))),as.vector(t(assay(input_data[rownames(input_data) == y,input_data$ImageNumber == imagenumber],select_assay))),
                        xlab = paste("Imagenumber:",imagenumber , "Marker:" , x, sep = " "),
                        ylab = y,
                        xlim = xlim,
                        ylim = ylim,
                        main=(paste("Image number:", imagenumber,sep =" ")),
                        cor = TRUE,
                        method = "spearman")
      }

}
```


# load data

```{r load data}
sce = readRDS(file = "~/bbvolume/Data/Dermatology/covid_skin_rash/Rout/data/sce_protein.rds")

# here we exclude two samples from the downstream analysis. DRESS_4 is characterized by very thick layer of keratinocytes and does not contain any dermis. NHS_1 is a HC samples that contains my T cells and was a normal control sample that was taken to close to an area of disease.
sce <- sce[,! sce$sample %in% c("DRESS_4","NHS_1")]

colour_vector <- metadata(sce)$colour_vector$layer_1

image_mat <- as.data.frame(read.csv(file = "~/bbvolume/Data/Dermatology/covid_skin_rash/processed/cpout/Image.csv",stringsAsFactors = FALSE))

sample_colors <- c("#F8766D", "#7CAE00","#00BFC4","#C77CFF")
names(sample_colors) <- c("Cov","DRESS","MDR","NHS")
```

# rename samples

```{r}
sce[,sce$sample == "MDR_1_lefstide"]$sample <- "MDR_1"
sce[,sce$sample == "Cov_1_top"]$sample <- "Cov_1"
```

```{r, fig.height=10,fig.width=10}
pdf("~/bbvolume/Data/Dermatology/covid_skin_rash/Rout/plots/cellfractions_classified.pdf",height = 10,width = 15)
plotCellFracGroups(sce,CellClass = "celltype_classified",color_by = "sample_type")
dev.off()
```

```{r}
cur_df <- data.frame(ImageNumber = sce$ImageNumber,
                     celltype = sce$celltype_classified,
                     sample = sce$sample,
                     sample_type = sce$sample_type)

cur_df <- reshape2::dcast(data = cur_df,formula = "ImageNumber + sample + sample_type   ~ celltype")

# calculate the fracions of each celltype per image
freqs_wide <- as.data.frame(t(scale(t(as.matrix(cur_df[,-c(1,2,3)])), center = FALSE, 
               scale = rowSums(cur_df[,-c(1,2,3)]))))

freqs_wide$ImageNumber <- cur_df$ImageNumber
freqs_wide$sample <- cur_df$sample
freqs_wide$sample_type <- cur_df$sample_type
freqs_wide$celltype <- cur_df$celltype

freqs_long <- reshape2::melt(freqs_wide,id.vars = c("ImageNumber","sample", "sample_type"),measure.vars=unique(sce$celltype_classified))

freqs_long$celltype <- sapply(as.vector(freqs_long$variable),function(x){strsplit(x,"_",)[[1]][1]})
```


```{r}
immune_celltypes <- c("Tcytotoxic","Thelper","Macrophages","Neutrophil","pDC","Bcell","Langerhans", "TcytotoxicMyeloid","ThelperMyeloid")

vasculature_cells <- "Vasculature"
```


```{r T cytotoxic cluster fractions per sample_type, fig.height=16,fig.width=10}

immune_celltype_fractions <- freqs_long %>%
  filter(celltype %in%  immune_celltypes) %>%
  arrange(celltype) %>%
  ggplot()+
  geom_boxplot(aes(x=celltype,y=value, fill=sample_type))+
  geom_jitter(aes(x=celltype,y=value, fill=sample_type), position = position_jitterdodge(0.05), lwd=0.6)+
  theme(axis.text.x = element_text(angle=90))+
  theme_bw()+
  theme(axis.text.x = element_text(angle=90))

pdf(file = "~/bbvolume/Data/Dermatology/covid_skin_rash/Rout/plots/cellfractions_classified_immune.pdf",width = 12,height = 10)
immune_celltype_fractions
dev.off()
```

```{r, fig.height=8,fig.width=10}
celltype_fractions_all <- freqs_long %>%
  arrange(celltype) %>%
  ggplot()+
  geom_boxplot(aes(x=celltype,y=value, fill=sample_type))+
  geom_jitter(aes(x=celltype,y=value, fill=sample_type), position = position_jitterdodge(0.05), lwd=0.6)+
  theme(axis.text.x = element_text(angle=90))+
  theme_bw()+
  theme(axis.text.x = element_text(angle=90))+
  scale_fill_manual(values = sample_colors)

pdf("~/bbvolume/Data/Dermatology/covid_skin_rash/Rout/plots/celltype_fractions_all.pdf", width = 12, height = 8)
celltype_fractions_all
dev.off()
```



## clustered celltypes classified boxplot per celltype

```{r generate celltype cluster fractions, message=FALSE}
# calculate the fractions of all celltype
cur_df <- data.frame(ImageNumber = sce$ImageNumber,
                     celltype_clustering_FS = sce$celltype_classified_clustering_FS,
                     celltype = sce$celltype_classified,
                     sample = sce$sample,
                     sample_type = sce$sample_type)

cur_df <- reshape2::dcast(data = cur_df,formula = "ImageNumber + sample + sample_type   ~ celltype_clustering_FS")

# calculate the fracions of each celltype per image
freqs_wide <- as.data.frame(t(scale(t(as.matrix(cur_df[,-c(1,2,3)])), center = FALSE, 
               scale = rowSums(cur_df[,-c(1,2,3)]))))

freqs_wide$ImageNumber <- cur_df$ImageNumber
freqs_wide$sample <- cur_df$sample
freqs_wide$sample_type <- cur_df$sample_type
freqs_wide$celltype <- cur_df$celltype

freqs_long <- reshape2::melt(freqs_wide,id.vars = c("ImageNumber","sample", "sample_type"),measure.vars=unique(sce$celltype_classified_clustering_FS), stringAsFactor= FALSE)

freqs_long$celltype <- sapply(as.vector(freqs_long$variable),function(x){strsplit(x,"_",)[[1]][1]})
freqs_long$cluster <- sapply(as.vector(freqs_long$variable),function(x){strsplit(x,"_",)[[1]][2]})

```


```{r clustered celltypes fractions per celltype across indications}

for (i in unique(freqs_long$celltype)) {
p <- freqs_long %>%
  filter(celltype == i) %>%
  arrange(cluster,decreasing = FALSE) %>%
  mutate(celltype = paste(i,cluster,sep="_")) %>%
  ggplot()+
    geom_boxplot(aes(x=celltype,y=value, fill=sample_type))+
    geom_jitter(aes(x=celltype,y=value, fill=sample_type), position = position_jitterdodge(0.05), lwd=0.6)+
      theme(axis.text.x = element_text(angle = 45, hjust = 1),
            axis.title.x = element_blank(),
            panel.background = element_blank(),
            panel.grid.minor=element_blank(),
            panel.grid.major.x=element_blank(),
            panel.grid.major.y=element_line(color="grey", size=.3))+
  scale_fill_manual(values = sample_colors)
pdf(file = paste0("~/bbvolume/Data/Dermatology/covid_skin_rash/Rout/plots/",i,"_classified_clustered.pdf"))
print(p)
dev.off()
}
```

```{r significance testing clustered celltypes classified, include=FALSE}
plot_dat <- freqs_long %>%
  filter(variable %in%  c("Tcytotoxic_3")) %>%
  arrange(celltype)

  ggboxplot(plot_dat,x="sample_type",y="value", color = "variable")+
  theme(axis.text.x = element_text(angle=90))+
  theme_bw()+
  theme(axis.text.x = element_text(angle=90))+
  stat_compare_means(comparisons = list( c("Cov", "DRESS"), c("Cov", "MDR"), c("Cov", "NHS")))
```


## Marker expression of celltypes classified

```{r heatmap marker expresion celltypes classified with annotation, fig.height=12,fig.width=8}
agr_sce <- aggregateAcrossCells(sce, ids = colData(sce)[,c("celltype_classified")], 
                                statistics = "mean")
assay(agr_sce, "asinh") <- asinh(counts(agr_sce))


plot_mat <- t(scale(t(assay(agr_sce,"asinh"))))

annotation_dat <- DataFrame(celltype = sce$celltype_classified)

annotation_dat <- as.data.frame(table(colData(sce)$celltype_classified))
  

row_ha = HeatmapAnnotation(size = anno_barplot(annotation_dat$Freq), celltype = annotation_dat$Var1,col=list(celltype = colour_vector))

pdf("~/bbvolume/Data/Dermatology/covid_skin_rash/Rout/plots/Marker_expression_celltype_classified_heatmap_with_annotation.pdf",width = 10,height = 10)
Heatmap(plot_mat,top_annotation = row_ha,height = unit(15,"cm"),width = unit(10,"cm"),cluster_columns = FALSE,
        col = colorRamp2(c(-3, 0, 3), c("dark blue", "white", "dark red")),heatmap_legend_param = list(at = c(-3, 0, 3)))
dev.off()
```

## Marker expression of celltypes classified clustered


```{r heatmap marker expresion celltypes classified clustered with annotation, fig.height=12,fig.width=12}
agr_sce <- aggregateAcrossCells(sce, ids = colData(sce)[,c("celltype_classified_clustering_FS")], 
                               statistics = "mean")
assay(agr_sce, "asinh") <- asinh(counts(agr_sce))


plot_mat <- t(scale(t(assay(agr_sce,"asinh"))))


annotation_dat <- as.data.frame(table(colData(sce)$celltype_classified_clustering_FS))
annotation_dat$celltype <- sapply(X =annotation_dat$Var1,FUN = function(X){str_split(X,pattern = "_")[[1]][1]})

row_ha = HeatmapAnnotation(size = anno_barplot(annotation_dat$Freq), celltype = annotation_dat$celltype,col=list(celltype=colour_vector))

pdf("~/bbvolume/Data/Dermatology/covid_skin_rash/Rout/plots/Marker_expression_celltype_classified_clustered_heatmap_with_annotation.pdf",width = 15,height = 10)
Heatmap(plot_mat,top_annotation = row_ha,height = unit(15,"cm"),width = unit(20,"cm"),cluster_columns = FALSE,
        col = colorRamp2(c(-3, 0, 3), c("dark blue", "white", "dark red")),heatmap_legend_param = list(at = c(-3, 0, 3)))
dev.off()
```

## Heatmap celltype classified clustered

```{r heatmap marker expresion celltypes classified clustered, fig.height=8,fig.width=12}
agr_sce <- aggregateAcrossCells(sce, ids = colData(sce)[,c("celltype_classified_clustering_FS")], 
                                statistics = "mean")
assay(agr_sce, "asinh") <- asinh(counts(agr_sce))


# Non-scaled
dittoHeatmap(agr_sce, assay = "asinh",
            annot.by = c("celltype_classified_clustering_FS"), 
            order.by = "celltype_classified_clustering_FS", cluster_rows = FALSE,
            scale = "none", heatmap.colors = viridis(100), 
            annotation_colors = list(celltype_classified = colour_vector))+theme(legend.position = "bottom")

# Centered and scaled
p <- dittoHeatmap(agr_sce, assay = "asinh",
            annot.by = c("celltype_classified"),
            annotation_colors = list(celltype_classified = colour_vector),
            heatmap.colors = colorRampPalette(c("dark blue", "white", "dark red"))(100),
            breaks = seq(-3,3, length.out = 101),show_colnames = TRUE)
  theme(axis.title.x = element_text("test"))

pdf(file = "~/bbvolume/Data/Dermatology/covid_skin_rash/Rout/plots/Marker_expr_heatmap_celltypes_classified_clustered_dittoHeatmap.pdf", width = 15, height = 10)
p
dev.off()
```



## Marker expression of celltypes classified split by indication

```{r heatmap marker expresion celltypes classified by indication with annotation, fig.height=12,fig.width=12}
agr_sce <- aggregateAcrossCells(sce, ids = colData(sce)[,c("sample_type","celltype_classified")], 
                                statistics = "mean")
assay(agr_sce, "asinh") <- asinh(counts(agr_sce))


plot_mat <- t(scale(t(assay(agr_sce,"asinh"))))
colnames(plot_mat) <- paste0(agr_sce$celltype_classified,"_",agr_sce$sample_type)


plot_mat <- plot_mat[,order(colnames(plot_mat))]

annotation_dat <- DataFrame(size = agr_sce$ncells,celltype = paste0(agr_sce$celltype_classified,"_",agr_sce$sample_type),sample_type = agr_sce$sample_type,celltype_cols =agr_sce$celltype_classified)

annotation_dat <- annotation_dat[order(annotation_dat$celltype),]

row_ha = HeatmapAnnotation(size = anno_barplot(annotation_dat$size),sample_type = annotation_dat$sample_type,celltype=annotation_dat$celltype_cols,col=list(celltype = colour_vector,sample_type = c("Cov"="#1B9E77","DRESS"="#D95F02","MDR" ="#7570B3", "NHS"= "#E7298A")))

pdf("~/bbvolume/Data/Dermatology/covid_skin_rash/Rout/plots/Marker_expression_celltype_classified__by_indication_heatmap_with_annotation.pdf",width = 15,height = 10)
Heatmap(plot_mat,top_annotation = row_ha,height = unit(15,"cm"),width = unit(20,"cm"),cluster_columns = FALSE,
        col = colorRamp2(c(-3, 0, 3), c("dark blue", "white", "dark red")),heatmap_legend_param = list(at = c(-3, 0, 3)))
dev.off()
```

```{r heatmap marker expresion celltypes classified by sample_type, fig.height=8,fig.width=10}
agr_sce <- aggregateAcrossCells(sce, ids = colData(sce)[,c("sample_type","celltype_classified")], 
                                statistics = "mean")
assay(agr_sce, "asinh") <- asinh(counts(agr_sce))

colnames(agr_sce) <- paste0(agr_sce$sample_type, "_", agr_sce$celltype_classified)

# Non-scaled
dittoHeatmap(agr_sce, assay = "asinh",
            annot.by = c("sample_type","celltype_classified"), 
            order.by = "celltype_classified", cluster_rows = FALSE,
            scale = "none", heatmap.colors = viridis(100), 
            annotation_colors = list(celltype_classified = colour_vector))

# Centered and scaled
p <- dittoHeatmap(agr_sce, assay = "asinh",
            annot.by = c("sample_type","celltype_classified"),
            order.by = "celltype_classified",
            annotation_colors = list(celltype_classified = colour_vector),
            heatmap.colors = colorRampPalette(c("dark blue", "white", "dark red"))(100),
            breaks = seq(-4,4, length.out = 101))
pdf(file = "~/bbvolume/Data/Dermatology/covid_skin_rash/Rout/plots/Marker_expr_heatmap_celltypes_classified_by_indication.pdf", width = 10, height = 8)
p
dev.off()
```


```{r UMAP, fig.height=20,fig.width=10}
p1 <- dittoDimPlot(sce,var = "celltype_classified",color.panel = colour_vector, size = 0.5)

p2 <- dittoDimPlot(sce,var = "sample_type", size = 0.5, color.panel = sample_colors)

UMAP_plot <- plot_grid(p1,p2, ncol=1)

jpeg("~/bbvolume/Data/Dermatology/covid_skin_rash/Rout/plots/UMAP.jpeg",width = 2000,height = 3000,units = "px", res = 300)
plot_grid(p1,p2, ncol=1)
dev.off()
```

## Fig. 2

```{r}
UMAP_plot
immune_celltype_fractions
```

```{r significance testing, include=FALSE}
cur_df <- data.frame(ImageNumber = sce$ImageNumber,
                     celltype = sce$celltype_classified_clustering_FS,
                     sample = sce$sample,
                     sample_type = sce$sample_type)

cur_df <- reshape2::dcast(data = cur_df,formula = "ImageNumber + sample + sample_type   ~ celltype")

# calculate the fracions of each celltype per image
freqs_wide <- as.data.frame(t(scale(t(as.matrix(cur_df[,-c(1,2,3)])), center = FALSE, 
               scale = rowSums(cur_df[,-c(1,2,3)]))))

freqs_wide$ImageNumber <- cur_df$ImageNumber
freqs_wide$sample <- cur_df$sample
freqs_wide$sample_type <- cur_df$sample_type
freqs_wide$celltype <- cur_df$celltype

freqs_long <- reshape2::melt(freqs_wide,id.vars = c("ImageNumber","sample", "sample_type"),measure.vars=unique(sce$celltype_classified_clustering_FS))

#freqs_long$celltype <- sapply(as.vector(freqs_long$variable),function(x){strsplit(x,"_",)[[1]][1]})

plot_dat <- freqs_long %>%
  filter(variable %in%  c("Macrophages_3")) %>%
  arrange(variable)
  
ggboxplot(plot_dat,x="sample_type",y="value")+
  theme(axis.text.x = element_text(angle=90))+
  theme_bw()+
  theme(axis.text.x = element_text(angle=90))+
  stat_compare_means(comparisons = list( c("Cov", "DRESS"),c("Cov","MDR"), c("Cov", "NHS")))

#
dat1 <- freqs_long[which(freqs_long$variable == "Macrophages_1" & freqs_long$sample_type == "Cov"),]$value

dat2 <- freqs_long[which(freqs_long$variable == "Macrophages_1" & freqs_long$sample_type == "DRESS"),]$value

t.test(x = dat1,y = dat2)
  
```
