---
title: "cellular interactions"
author: "daniels"
date: "19 5 2020"
output: html_document
---

here we use the interaction counts as calculated by CellProfiler to plot the interactions amongst cell types in the different indications.

CellProfiler counts interactions between cells when they are maximally 8 pixels apart from each other. The data is then outputted in a "relationship" table which lists the IDs of cells and all IDs of cells neighbouring a cell.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

## load packages

```{r load packages, echo=FALSE, message = FALSE}
sapply(list.files("~/bbvolume/Git/imcRtools/R/", full.names = TRUE), source)

library(LSD)
library(data.table)
library(dplyr)
library(magrittr)
library(dtplyr)
library(parallel)
library(gplots)
library(SingleCellExperiment)
library(ggplot2)
library(Rphenograph)
library(bbRtools)
library(neighbouRhood)
library(RColorBrewer)
library(igraph)
library(circlize)
```

# load data

```{r path to data}
fn_cells <- "~/bbvolume/Data/Dermatology/covid_skin_rash//processed/cpout/cell.csv"

fn_relationship <- "~/bbvolume/Data/Dermatology/covid_skin_rash/processed/cpout/Object relationships.csv"
```


```{r load data}
sce = readRDS(file = "~/bbvolume/Data/Dermatology/covid_skin_rash/Rout/data/sce_protein.rds")

dat_relation <- fread(fn_relationship)

sce[,sce$sample == "MDR_1_lefstide"]$sample <- "MDR_1"
sce[,sce$sample == "Cov_1_top"]$sample <- "Cov_1"

sce <- sce[,! sce$sample %in% c("DRESS_4","NHS_1")]
```

## Choord plots per image
here we use choord plots to display the interaction counts among cell-types in individual images

```{r, fig.height=20, fig.width=20}

pdf("~/bbvolume/Data/Dermatology/covid_skin_rash/Rout/plots/chordDiagram_all_samples.pdf",width = 20,height = 20)

layout(matrix(c(1:4,c(5:7,7),c(8,9,10,11),c(12,13,14,14)),byrow = TRUE,nc=4))

#layout(matrix(1:16,nc=4,byrow=TRUE))

for (i in unique(sce$ImageNumber)){

  cur_sce <- sce[,sce$ImageNumber == i]
  
  title_name <- unique(cur_sce$sample)

  celltype_dat_First <- DataFrame(FirstObjectNumber = cur_sce$cellID,
                           celltype = cur_sce$celltype_classified)
  celltype_dat_Second <- DataFrame(SecondObjectNumber = cur_sce$cellID,
                           celltype = cur_sce$celltype_classified)


  cur_dat_relation <- as.data.frame(dat_relation) %>%
    filter(`First Image Number` == i) %>%
    select("First Object Number" ,"Second Object Number") %>%
    mutate(FirstObjectNumber = paste0(i,"_",`First Object Number`),SecondObjectNumber = paste0(i,"_",`Second Object Number`)) %>%
    select(3:4)

  cur_dat_relation <- left_join(x=cur_dat_relation,y=celltype_dat_First,by="FirstObjectNumber",copy=TRUE)
  cur_dat_relation <- left_join(x=cur_dat_relation,y= celltype_dat_Second,by="SecondObjectNumber",copy=TRUE)
  
  adj_list <- cur_dat_relation %>%
    select(3:4) %>%
    filter(celltype.x != "Keratinocyte1", celltype.y != "Keratinocyte1") %>%
    dplyr::count(celltype.x,celltype.y)

chordDiagram(adj_list, grid.col = metadata(sce)$colour_vectors$layer_1, annotationTrack = "grid", 
    preAllocateTracks = list(track.height = max(strwidth(unlist(dimnames(adj_list))))))
for(si in get.all.sector.index()) {
    circos.axis(h = "top", labels.cex = 0.5, sector.index = si, track.index = 2)
}
circos.track(track.index = 1, panel.fun = function(x, y) {
    xlim = get.cell.meta.data("xlim")
    ylim = get.cell.meta.data("ylim")
    sector.name = get.cell.meta.data("sector.index")
    xplot = get.cell.meta.data("xplot")
    
    circos.text(CELL_META$xcenter, CELL_META$ylim[1], CELL_META$sector.index, 
          facing = "clockwise", niceFacing = TRUE, adj = c(-0.15, 0.5),cex=1)
  }, bg.border = NA) # here set bg.border to NA is important
  title(main = title_name)
circos.clear()
}
#adj_matrix <- ftM2adjM(as.matrix(cur_dat_relation[,1:2]))
dev.off()

```
## chordDiagrams per indication

```{r, fig.height=6, fig.width=25}

pdf("~/bbvolume/Data/Dermatology/covid_skin_rash/Rout/plots/chordDiagram_by_indication.pdf",width = 20,height =5.5)

layout(matrix(1:4,1,4))

for (i in unique(sce$sample_type)){

  cur_sce <- sce[,sce$sample_type == i]
  
  title_name <- unique(cur_sce$sample_type)

  celltype_dat_First <- DataFrame(FirstObjectNumber = cur_sce$cellID,
                           celltype = cur_sce$celltype_classified)
  celltype_dat_Second <- DataFrame(SecondObjectNumber = cur_sce$cellID,
                           celltype = cur_sce$celltype_classified)
  
  sample_type_image <- unique(cur_sce$ImageNumber)


  cur_dat_relation <- as.data.frame(dat_relation) %>%
    filter(`First Image Number` %in% sample_type_image) %>%
    select("First Object Number" ,"Second Object Number", "First Image Number") %>%
    mutate(FirstObjectNumber = paste0(`First Image Number`,"_",`First Object Number`),SecondObjectNumber = paste0(`First Image Number`,"_",`Second Object Number`)) %>%
    select(4:5)

  cur_dat_relation <- left_join(x=cur_dat_relation,y=celltype_dat_First,by="FirstObjectNumber",copy=TRUE)
  cur_dat_relation <- left_join(x=cur_dat_relation,y= celltype_dat_Second,by="SecondObjectNumber",copy=TRUE)
  
  adj_list <- cur_dat_relation %>%
    select(3:4) %>%
    filter(celltype.x != "Keratinocyte1", celltype.y != "Keratinocyte1") %>%
    dplyr::count(celltype.x,celltype.y)


chordDiagram(adj_list, grid.col = metadata(sce)$colour_vectors$layer_1, annotationTrack = "grid", 
    preAllocateTracks = list(track.height = max(strwidth(unlist(dimnames(adj_list))))))
for(si in get.all.sector.index()) {
    circos.axis(h = "top", labels.cex = 0.5, sector.index = si, track.index = 2)
}
circos.track(track.index = 1, panel.fun = function(x, y) {
    xlim = get.cell.meta.data("xlim")
    ylim = get.cell.meta.data("ylim")
    sector.name = get.cell.meta.data("sector.index")
    xplot = get.cell.meta.data("xplot")
    
    circos.text(CELL_META$xcenter, CELL_META$ylim[1], CELL_META$sector.index, 
          facing = "clockwise", niceFacing = TRUE, adj = c(-0.15, 0.5),cex=1)
  }, bg.border = NA) # here set bg.border to NA is important
  title(main = title_name)
circos.clear()
}
#adj_matrix <- ftM2adjM(as.matrix(cur_dat_relation[,1:2]))
dev.off()

```
## chordDiagrams per indication for all celltypes

```{r, fig.height=6, fig.width=25}
pdf("~/bbvolume/Data/Dermatology/covid_skin_rash/Rout/plots/chordDiagram_by_indication_all_celltypes.pdf",width = 20,height =5.5)

layout(matrix(1:4,1,4))

for (i in unique(sce$sample_type)){

  cur_sce <- sce[,sce$sample_type == i]
  
  title_name <- unique(cur_sce$sample_type)

  celltype_dat_First <- DataFrame(FirstObjectNumber = cur_sce$cellID,
                           celltype = cur_sce$celltype_classified)
  celltype_dat_Second <- DataFrame(SecondObjectNumber = cur_sce$cellID,
                           celltype = cur_sce$celltype_classified)
  
  sample_type_image <- unique(cur_sce$ImageNumber)


  cur_dat_relation <- as.data.frame(dat_relation) %>%
    filter(`First Image Number` %in% sample_type_image) %>%
    select("First Object Number" ,"Second Object Number", "First Image Number") %>%
    mutate(FirstObjectNumber = paste0(`First Image Number`,"_",`First Object Number`),SecondObjectNumber = paste0(`First Image Number`,"_",`Second Object Number`)) %>%
    select(4:5)

  cur_dat_relation <- left_join(x=cur_dat_relation,y=celltype_dat_First,by="FirstObjectNumber",copy=TRUE)
  cur_dat_relation <- left_join(x=cur_dat_relation,y= celltype_dat_Second,by="SecondObjectNumber",copy=TRUE)
  
  adj_list <- cur_dat_relation %>%
    select(3:4) %>%
    dplyr::count(celltype.x,celltype.y)


chordDiagram(adj_list, grid.col = metadata(sce)$colour_vectors$layer_1, annotationTrack = "grid", 
    preAllocateTracks = list(track.height = max(strwidth(unlist(dimnames(adj_list))))))
for(si in get.all.sector.index()) {
    circos.axis(h = "top", labels.cex = 0.5, sector.index = si, track.index = 2)
}
circos.track(track.index = 1, panel.fun = function(x, y) {
    xlim = get.cell.meta.data("xlim")
    ylim = get.cell.meta.data("ylim")
    sector.name = get.cell.meta.data("sector.index")
    xplot = get.cell.meta.data("xplot")
    
    circos.text(CELL_META$xcenter, CELL_META$ylim[1], CELL_META$sector.index, 
          facing = "clockwise", niceFacing = TRUE, adj = c(-0.15, 0.5),cex=1)
  }, bg.border = NA) # here set bg.border to NA is important
  title(main = title_name)
circos.clear()
}
#adj_matrix <- ftM2adjM(as.matrix(cur_dat_relation[,1:2]))
dev.off()

```

## chordDiagrams per indication without outlier samples

```{r, fig.height=6, fig.width=25}
sce <- sce[,! sce$sample %in% c("DRESS_4","NHS_1", "Cov_1","MDR_1")]

pdf("~/bbvolume/Data/Dermatology/covid_skin_rash/Rout/plots/chordDiagram_by_indication_no_outliers.pdf",width = 20,height =5.5)

layout(matrix(1:4,1,4))

for (i in unique(sce$sample_type)){

  cur_sce <- sce[,sce$sample_type == i]
  
  title_name <- unique(cur_sce$sample_type)

  celltype_dat_First <- DataFrame(FirstObjectNumber = cur_sce$cellID,
                           celltype = cur_sce$celltype_classified)
  celltype_dat_Second <- DataFrame(SecondObjectNumber = cur_sce$cellID,
                           celltype = cur_sce$celltype_classified)
  
  sample_type_image <- unique(cur_sce$ImageNumber)


  cur_dat_relation <- as.data.frame(dat_relation) %>%
    filter(`First Image Number` %in% sample_type_image) %>%
    select("First Object Number" ,"Second Object Number", "First Image Number") %>%
    mutate(FirstObjectNumber = paste0(`First Image Number`,"_",`First Object Number`),SecondObjectNumber = paste0(`First Image Number`,"_",`Second Object Number`)) %>%
    select(4:5)

  cur_dat_relation <- left_join(x=cur_dat_relation,y=celltype_dat_First,by="FirstObjectNumber",copy=TRUE)
  cur_dat_relation <- left_join(x=cur_dat_relation,y= celltype_dat_Second,by="SecondObjectNumber",copy=TRUE)
  
  adj_list <- cur_dat_relation %>%
    select(3:4) %>%
    filter(celltype.x != "Keratinocyte1", celltype.y != "Keratinocyte1") %>%
    dplyr::count(celltype.x,celltype.y)


chordDiagram(adj_list, grid.col = metadata(sce)$colour_vectors$layer_1, annotationTrack = "grid", 
    preAllocateTracks = list(track.height = max(strwidth(unlist(dimnames(adj_list))))))
for(si in get.all.sector.index()) {
    circos.axis(h = "top", labels.cex = 0.5, sector.index = si, track.index = 2)
}
circos.track(track.index = 1, panel.fun = function(x, y) {
    xlim = get.cell.meta.data("xlim")
    ylim = get.cell.meta.data("ylim")
    sector.name = get.cell.meta.data("sector.index")
    xplot = get.cell.meta.data("xplot")
    
    circos.text(CELL_META$xcenter, CELL_META$ylim[1], CELL_META$sector.index, 
          facing = "clockwise", niceFacing = TRUE, adj = c(-0.15, 0.5),cex=1)
  }, bg.border = NA) # here set bg.border to NA is important
  title(main = title_name)
circos.clear()
}
#adj_matrix <- ftM2adjM(as.matrix(cur_dat_relation[,1:2]))
dev.off()

```